% =============================================================================
% Lecture 07: Introduction to Neural Networks (PyTorch)
% BSAD 8310: Business Forecasting
% University of Nebraska at Omaha
% =============================================================================

\documentclass[aspectratio=169, 11pt]{beamer}

\input{header}

% ---- Lecture metadata --------------------------------------------------------
\title{Introduction to Neural Networks}
\subtitle{BSAD 8310: Business Forecasting --- Lecture 7}
\author{Department of Economics}
\institute{University of Nebraska at Omaha}
\date{Spring 2026}

% =============================================================================
\begin{document}
% =============================================================================

\begin{frame}
  \titlepage
\end{frame}

% ---- Outline ----------------------------------------------------------------
\begin{frame}{Lecture Outline}
  \tableofcontents
\end{frame}

% =============================================================================
\section{From Trees to Neural Networks}
% =============================================================================

\sectionslide{From Trees to Neural Networks}{%
  Tree-based methods are powerful but limited by their piecewise-constant
  predictions. Neural networks learn smooth, continuous functions of
  arbitrary complexity.}

% --- Slide: Motivation -------------------------------------------------------
\begin{frame}{Why Go Beyond Trees?}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Tree-based models (L04--L06):}
      \begin{itemize}\small
        \item Excellent for tabular data with many features
        \item Each row treated \textbf{independently} ---
              no built-in notion of order or sequence
        \item Predictions are piecewise constant:
              cannot extrapolate beyond training range
        \item Feature engineering required for temporal patterns
              (lag columns, rolling means, etc.)
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \textbf{What neural networks add:}
      \begin{itemize}\small
        \item Learn \textbf{smooth, continuous} functions directly
        \item Flexible architecture: can be adapted
              for sequences, images, text
        \item Can learn hierarchical feature representations
              automatically
        \item PyTorch enables custom architectures and
              GPU-accelerated training
      \end{itemize}
    \end{column}
  \end{columns}
  \vspace{0.2cm}
  \begin{keybox}
    Neural networks are not always better than XGBoost on tabular data.
    They shine when:
    (1) the data is a sequence (time series, text),
    (2) raw inputs are images or audio, or
    (3) you have very large amounts of data ($n > 100\text{K}$).
  \end{keybox}
\end{frame}

% --- Slide: From regression to a neuron ---------------------------------------
\begin{frame}{From Linear Regression to a Single Neuron}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Linear regression} (Lecture 02):
      \[
        \hat{y} = \mathbf{w}^\top \mathbf{x} + b
      \]
      \textbf{One neuron} wraps the linear combination in a nonlinear
      \textbf{activation function} $\sigma(\cdot)$:
      \[
        a = \sigma(\mathbf{w}^\top \mathbf{x} + b)
      \]
      \begin{center}
        {\footnotesize
        \begin{tabular}{lll}
          \toprule
          \textbf{Activation} & \textbf{Formula} & \textbf{Typical use} \\
          \midrule
          ReLU    & $\max(0,z)$       & Hidden layers \\
          Sigmoid & $1/(1+e^{-z})$    & Binary output / gates \\
          Tanh    & $\tanh(z)$        & Sequence hidden states \\
          Linear  & $z$               & Regression output \\
          \bottomrule
        \end{tabular}
        }
      \end{center}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{definitionbox}{Why Non-Linearity?}
        Without $\sigma(\cdot)$, stacking layers is equivalent
        to a single linear model --- the composition of linear
        functions is still linear.

        Non-linearity enables a network to approximate
        \textbf{any continuous function} to arbitrary accuracy
        (universal approximation theorem).
      \end{definitionbox}
      \vspace{0.1cm}
      \begin{examplebox}{Business intuition}
        A neuron asks: ``Is this pattern present?''
        If yes (large positive $z$), activate.
        If no, stay off. ReLU makes this literal.
      \end{examplebox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Feedforward Networks}
% =============================================================================

\sectionslide{Feedforward Networks}{%
  Stack layers of neurons. Each layer learns higher-level
  representations of the input.}

% --- Slide: Layers and forward pass ------------------------------------------
\begin{frame}{Feedforward Network: Layers and Forward Pass}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      A \textbf{feedforward network} (FFN / MLP) has:
      \begin{itemize}\small
        \item \textbf{Input layer:} raw features
              $\mathbf{x} \in \mathbb{R}^p$
        \item \textbf{Hidden layers:} $H$ layers, each applying
              $\mathbf{a}^{(\ell)} = \sigma(\mathbf{W}^{(\ell)}\mathbf{a}^{(\ell-1)} + \mathbf{b}^{(\ell)})$
        \item \textbf{Output layer:} linear (regression)
              or softmax (classification)
      \end{itemize}
      \vspace{0.2cm}
      \textbf{Notation:}
      \begin{itemize}\small
        \item $\mathbf{W}^{(\ell)}$: weight matrix for layer $\ell$
        \item $\mathbf{b}^{(\ell)}$: bias vector for layer $\ell$
        \item Width = number of neurons per layer
        \item Depth = number of hidden layers
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{keybox}
        \small
        \textbf{Practical sizing for forecasting:}
        \begin{itemize}\small
          \item 2--3 hidden layers
          \item 64--256 neurons per layer
          \item ReLU activations
          \item Dropout 0.2--0.5 for regularization
        \end{itemize}
        Wider/deeper networks need more data
        and more regularization.
      \end{keybox}
      \vspace{0.1cm}
      \begin{warningbox}
        \small Bigger is not always better.
        Start small, add capacity only if
        the validation error is still too high.
      \end{warningbox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Training: Loss and Backpropagation}
% =============================================================================

\sectionslide{Training: Loss and Backpropagation}{%
  How does the network learn? By computing the error,
  then propagating it backward to update weights.}

% --- Slide: Loss and SGD -----------------------------------------------------
\begin{frame}{Training a Network: Loss, Gradients, and SGD}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Step 1: Define a loss function}

      For regression: $\mathcal{L} = \frac{1}{n}\sum_i (y_i - \hat{y}_i)^2$
      \hfill (Mean Squared Error)

      \vspace{0.2cm}
      \textbf{Step 2: Backpropagation}

      Apply the chain rule to compute $\frac{\partial \mathcal{L}}{\partial \mathbf{W}^{(\ell)}}$
      for every weight matrix. PyTorch does this automatically.

      \vspace{0.2cm}
      \textbf{Step 3: Update weights (SGD step)}
      \[
        \mathbf{W} \leftarrow \mathbf{W} - \eta \frac{\partial \mathcal{L}}{\partial \mathbf{W}}
      \]
      where $\eta$ is the \textbf{learning rate}.
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{definitionbox}{Stochastic Gradient Descent}
        Instead of computing the gradient on all $n$ observations
        at once (slow), use a random \textbf{mini-batch} of size
        32--256 per step. One pass through all mini-batches
        is one \textbf{epoch}.
      \end{definitionbox}
      \vspace{0.1cm}
      \begin{keybox}
        \small
        \textbf{In practice, use Adam optimizer}
        instead of plain SGD. Adam adapts the learning rate
        per parameter and converges much faster.
        \texttt{torch.optim.Adam(lr=1e-3)}
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: Overfitting and regularization ------------------------------------
\begin{frame}{Preventing Overfitting: Dropout and Early Stopping}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \begin{definitionbox}{Dropout}
        During training, randomly set a fraction $p$
        of neuron outputs to zero on each forward pass.
        This prevents co-adaptation: neurons cannot
        rely on specific other neurons being present.

        At test time, all neurons are active
        and outputs are scaled by $(1-p)$.
      \end{definitionbox}
      \vspace{0.1cm}
      PyTorch: \texttt{nn.Dropout(p=0.3)}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{warningbox}
        \textbf{Early stopping} is essential.
        Monitor validation loss after every epoch.
        Stop training when validation loss stops
        improving for $k$ consecutive epochs (patience).
        Save the model at the best validation epoch.
      \end{warningbox}
      \vspace{0.1cm}
      \begin{keybox}
        \small
        \textbf{Standard regularization toolkit:}
        \begin{itemize}\small
          \item Dropout (p = 0.2--0.5)
          \item Weight decay in Adam
                (\texttt{weight\_decay=1e-4})
          \item Early stopping (patience = 10--20)
          \item Batch normalization (deeper nets)
        \end{itemize}
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{PyTorch Implementation}
% =============================================================================

\sectionslide{PyTorch Implementation}{%
  Dataset, DataLoader, model definition, and training loop.}

% --- Slide: PyTorch Dataset and DataLoader -----------------------------------
\begin{frame}{PyTorch: Dataset and DataLoader}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \begin{examplebox}{Custom Dataset}
        {\small
        \texttt{import torch}\\
        \texttt{from torch.utils.data import Dataset, DataLoader}\\[4pt]
        \texttt{class ForecastDataset(Dataset):}\\
        \texttt{  def \_\_init\_\_(self, X, y):}\\
        \texttt{    self.X = torch.tensor(X, dtype=torch.float32)}\\
        \texttt{    self.y = torch.tensor(y, dtype=torch.float32)}\\[4pt]
        \texttt{  def \_\_len\_\_(self):}\\
        \texttt{    return len(self.X)}\\[4pt]
        \texttt{  def \_\_getitem\_\_(self, idx):}\\
        \texttt{    return self.X[idx], self.y[idx]}\\[4pt]
        \texttt{train\_ds = ForecastDataset(X\_train, y\_train)}\\
        \texttt{train\_loader = DataLoader(}\\
        \texttt{  train\_ds, batch\_size=64, shuffle=True)}
        }
      \end{examplebox}
    \end{column}
    \begin{column}{0.44\textwidth}
      \textbf{Why Dataset and DataLoader?}
      \begin{itemize}\small
        \item \texttt{Dataset}: wraps your data and
              defines how to get one sample
        \item \texttt{DataLoader}: handles mini-batching,
              shuffling, and parallel loading
        \item Separates data handling from model logic ---
              easy to swap in different datasets
      \end{itemize}
      \vspace{0.2cm}
      \begin{warningbox}
        \small
        \textbf{Time series:} do NOT shuffle.
        Set \texttt{shuffle=False} and use
        \texttt{TimeSeriesSplit} splits on the
        underlying indices before creating the
        Dataset.
      \end{warningbox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: Model definition -------------------------------------------------
\begin{frame}{PyTorch: Defining the Model}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \begin{examplebox}{Feedforward Regressor}
        {\small
        \texttt{import torch.nn as nn}\\[4pt]
        \texttt{class FFNRegressor(nn.Module):}\\
        \texttt{  def \_\_init\_\_(self, in\_dim):}\\
        \texttt{    super().\_\_init\_\_()}\\
        \texttt{    self.net = nn.Sequential(}\\
        \texttt{      nn.Linear(in\_dim, 128),}\\
        \texttt{      nn.ReLU(),}\\
        \texttt{      nn.Dropout(0.3),}\\
        \texttt{      nn.Linear(128, 64),}\\
        \texttt{      nn.ReLU(),}\\
        \texttt{      nn.Dropout(0.3),}\\
        \texttt{      nn.Linear(64, 1)}\\
        \texttt{    )}\\[4pt]
        \texttt{  def forward(self, x):}\\
        \texttt{    return self.net(x).squeeze(-1)}\\[4pt]
        \texttt{model = FFNRegressor(in\_dim=X\_train.shape[1])}
        }
      \end{examplebox}
    \end{column}
    \begin{column}{0.44\textwidth}
      \textbf{Key PyTorch concepts:}
      \begin{itemize}\small
        \item \texttt{nn.Module}: base class for all models.
              Override \texttt{\_\_init\_\_} and \texttt{forward}.
        \item \texttt{nn.Sequential}: chains layers in order
        \item \texttt{nn.Linear(in, out)}: fully connected layer
              with $in \times out$ weights + bias
        \item \texttt{nn.ReLU()}: element-wise ReLU activation
        \item \texttt{nn.Dropout(p)}: dropout layer
      \end{itemize}
      \vspace{0.1cm}
      \begin{keybox}
        \small The \texttt{forward} method defines what
        happens on each pass through the network.
        PyTorch automatically computes gradients
        via \texttt{autograd}.
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: Training loop ----------------------------------------------------
\begin{frame}{PyTorch: The Training Loop}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \begin{examplebox}{Training Loop}
        {\small
        \texttt{optimizer = torch.optim.Adam(}\\
        \texttt{  model.parameters(), lr=1e-3)}\\
        \texttt{criterion = nn.MSELoss()}\\[4pt]
        \texttt{for epoch in range(100):}\\
        \texttt{  model.train()}\\
        \texttt{  for X\_batch, y\_batch in train\_loader:}\\
        \texttt{    optimizer.zero\_grad()}\\
        \texttt{    y\_hat = model(X\_batch)}\\
        \texttt{    loss = criterion(y\_hat, y\_batch)}\\
        \texttt{    loss.backward()}\\
        \texttt{    optimizer.step()}\\[4pt]
        \texttt{  \# Evaluate on validation set}\\
        \texttt{  model.eval()}\\
        \texttt{  with torch.no\_grad():}\\
        \texttt{    val\_hat = model(X\_val\_t)}\\
        \texttt{    val\_loss = criterion(val\_hat, y\_val\_t)}\\
        \texttt{  print(f"Epoch \{epoch\}: val={val\_loss:.2f}")}
        }
      \end{examplebox}
    \end{column}
    \begin{column}{0.44\textwidth}
      \textbf{Four required steps per mini-batch:}
      \begin{enumerate}\small
        \item \texttt{zero\_grad()}: clear accumulated gradients
        \item Forward pass: compute $\hat{y}$
        \item \texttt{loss.backward()}: compute gradients
        \item \texttt{optimizer.step()}: update weights
      \end{enumerate}
      \vspace{0.2cm}
      \begin{warningbox}
        \small Always call \texttt{model.eval()} before
        validation to disable dropout.
        Use \texttt{torch.no\_grad()} to skip
        gradient computation during evaluation
        (saves memory and time).
      \end{warningbox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Key Takeaways}
% =============================================================================

\sectionslide{Key Takeaways}{%
  Neural networks are flexible function approximators.
  PyTorch provides the building blocks; the training loop
  is always the same four steps.}

% --- Slide: Method comparison ------------------------------------------------
\begin{frame}{Neural Networks vs.\ Tree Methods}
  \begin{center}
    {\small
    \begin{tabular}{lllll}
      \toprule
      \textbf{Method} & \textbf{Handles sequences} & \textbf{Extrapolates} & \textbf{Tuning burden} & \textbf{Best for} \\
      \midrule
      Random Forest & No (needs lags) & No & Low & Tabular, fast baseline \\
      XGBoost & No (needs lags) & No & Medium & Tabular, best accuracy \\
      FFN (MLP) & Partially & Yes & Medium & Tabular + smooth functions \\
      RNN / LSTM & Yes & Yes & High & Sequences, time series \\
      \midrule
      \multicolumn{5}{l}{\footnotesize Lectures 08--09 extend FFN to sequences and images.} \\
      \bottomrule
    \end{tabular}
    }
  \end{center}
  \begin{keybox}
    \small
    \textbf{When to use an FFN for forecasting:}
    \begin{itemize}\small
      \item You have engineered lag and rolling features (same as XGBoost)
      \item You want a smooth, continuous forecast function
      \item Baseline: XGBoost usually wins on small-to-medium tabular data;
            FFN may win with larger data or more complex patterns
    \end{itemize}
  \end{keybox}
\end{frame}

% --- Slide: Key takeaways ---------------------------------------------------
\begin{frame}{Lecture 7: Key Takeaways}
  \begin{keybox}
    {\small
    \begin{enumerate}
      \item A \textbf{neuron} combines inputs linearly, then applies a nonlinear
            \textbf{activation function} (ReLU, sigmoid, tanh). Without
            non-linearity, stacking layers is equivalent to one linear layer.

      \item A \textbf{feedforward network} stacks layers of neurons.
            Each layer learns higher-level representations.
            Depth adds capacity; width adds breadth.

      \item \textbf{Training} is iterative: compute loss, backpropagate gradients,
            update weights with Adam. Use mini-batches for efficiency.

      \item The \textbf{four mandatory steps} per mini-batch in PyTorch:
            \texttt{zero\_grad()} $\to$ forward pass $\to$
            \texttt{loss.backward()} $\to$ \texttt{optimizer.step()}.

      \item Prevent overfitting with \textbf{dropout}, \textbf{weight decay},
            and \textbf{early stopping} on a held-out validation set.

      \item For time series, do not shuffle the DataLoader.
            Use \textbf{TimeSeriesSplit} to create train/val splits.
    \end{enumerate}
    }
  \end{keybox}
  \muted{\itshape Next: CNN Architectures --- convolutional filters, pooling,
  and residual connections for structured data.}
\end{frame}

% --- References ---------------------------------------------------------------
\begin{frame}[allowframebreaks]{References}
  \printbibliography[heading=none]
\end{frame}

% =============================================================================
\end{document}
% =============================================================================
