% =============================================================================
% Lecture 09: RNNs, LSTMs & Transformers
% BSAD 8310: Business Forecasting
% University of Nebraska at Omaha
% =============================================================================

\documentclass[aspectratio=169, 11pt]{beamer}

\input{header}

% ---- Lecture metadata --------------------------------------------------------
\title{RNNs, LSTMs \& Transformers}
\subtitle{BSAD 8310: Business Forecasting --- Lecture 9}
\author{Department of Economics}
\institute{University of Nebraska at Omaha}
\date{Spring 2026}

% =============================================================================
\begin{document}
% =============================================================================

\begin{frame}
  \titlepage
\end{frame}

% ---- Outline ----------------------------------------------------------------
\begin{frame}{Lecture Outline}
  \tableofcontents
\end{frame}

% =============================================================================
\section{Recurrent Neural Networks}
% =============================================================================

\sectionslide{Recurrent Neural Networks}{%
  Process sequences step-by-step, maintaining a hidden state
  that carries information across time.}

% --- Slide: RNN motivation ---------------------------------------------------
\begin{frame}{Why Recurrent Networks?}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{The limitation of FFN and CNN:}
      \begin{itemize}\small
        \item FFN: each row processed independently ---
              temporal order ignored unless encoded as lag features
        \item 1D CNN: local pattern detection within a fixed window ---
              long-range dependencies require many stacked layers
      \end{itemize}
      \vspace{0.2cm}
      \textbf{What an RNN adds:}
      \begin{itemize}\small
        \item A \textbf{hidden state} $\mathbf{h}_t$ passed from step to step
        \item At each time step, the network updates its memory
              based on the new input \emph{and} what it remembered
        \item Theoretically, $\mathbf{h}_t$ encodes the entire history
              $x_1, x_2, \ldots, x_t$
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{definitionbox}{Vanilla RNN}
        At each step $t$:
        \[
          \mathbf{h}_t = \tanh(\mathbf{W}_h \mathbf{h}_{t-1} + \mathbf{W}_x \mathbf{x}_t + \mathbf{b})
        \]
        \[
          \hat{y}_t = \mathbf{W}_y \mathbf{h}_t + \mathbf{b}_y
        \]
        Same weights $\mathbf{W}_h, \mathbf{W}_x$ used at every step.
      \end{definitionbox}
      \begin{warningbox}
        \small \textbf{Vanishing gradient problem:}
        gradients from distant past steps shrink exponentially
        during backpropagation through time (BPTT).
        In practice, vanilla RNNs forget after $\sim$10 steps.
      \end{warningbox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{LSTM: Long Short-Term Memory}
% =============================================================================

\sectionslide{LSTM: Long Short-Term Memory}{%
  Gates control what to remember, what to forget,
  and what to output --- solving the vanishing gradient problem.}

% --- Slide: LSTM gates -------------------------------------------------------
\begin{frame}{LSTM: Three Gates and Cell State}
  \begin{columns}[T]
    \begin{column}{0.55\textwidth}
      An LSTM maintains two states: the \textbf{cell state} $\mathbf{c}_t$
      (long-term memory) and the \textbf{hidden state} $\mathbf{h}_t$ (output).

      \vspace{0.2cm}
      \textbf{Three gates + cell update per step:}
      \begin{enumerate}\small
        \item \textbf{Forget gate:} $\mathbf{f}_t = \sigma(\mathbf{W}_f[\mathbf{h}_{t-1}, \mathbf{x}_t] + \mathbf{b}_f)$\\
              What fraction of the old cell state to keep?
        \item \textbf{Input gate:} $\mathbf{i}_t = \sigma(\mathbf{W}_i[\mathbf{h}_{t-1}, \mathbf{x}_t] + \mathbf{b}_i)$\\
              What new information to write to memory?
        \item \textbf{Cell update:} candidate $\tilde{\mathbf{c}}_t = \tanh(\mathbf{W}_c[\mathbf{h}_{t-1}, \mathbf{x}_t] + \mathbf{b}_c)$\\
              then $\mathbf{c}_t = \mathbf{f}_t \odot \mathbf{c}_{t-1} + \mathbf{i}_t \odot \tilde{\mathbf{c}}_t$
        \item \textbf{Output gate:} $\mathbf{h}_t = \mathbf{o}_t \odot \tanh(\mathbf{c}_t)$\\
              What part of memory to expose as output?
      \end{enumerate}
    \end{column}
    \begin{column}{0.41\textwidth}
      \begin{keybox}
        \textbf{Why does LSTM solve vanishing gradients?}

        The cell state $\mathbf{c}_t$ is updated additively
        (via the forget gate). Gradients can flow backward
        through additive connections without shrinking ---
        similar to the residual connections in ResNet.
      \end{keybox}
      \vspace{0.1cm}
      \begin{examplebox}{Business analogy}
        The forget gate decides: ``Sales data from 3 years ago
        is not relevant; forget it.'' The input gate decides:
        ``This holiday spike is important; remember it.''
      \end{examplebox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: LSTM in PyTorch --------------------------------------------------
\begin{frame}{LSTM for Forecasting in PyTorch}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \begin{examplebox}{LSTM Regressor}
        {\small
        \texttt{class LSTMForecaster(nn.Module):}\\
        \texttt{  def \_\_init\_\_(self, in\_dim, hidden=128, layers=2):}\\
        \texttt{    super().\_\_init\_\_()}\\
        \texttt{    self.lstm = nn.LSTM(}\\
        \texttt{      in\_dim, hidden,}\\
        \texttt{      num\_layers=layers,}\\
        \texttt{      batch\_first=True,}\\
        \texttt{      dropout=0.2}\\
        \texttt{    )}\\
        \texttt{    self.fc = nn.Linear(hidden, 1)}\\[4pt]
        \texttt{  def forward(self, x):}\\
        \texttt{    \# x: (batch, seq\_len, in\_dim)}\\
        \texttt{    out, (h, c) = self.lstm(x)}\\
        \texttt{    \# Use last hidden state}\\
        \texttt{    return self.fc(h[-1]).squeeze(-1)}
        }
      \end{examplebox}
    \end{column}
    \begin{column}{0.44\textwidth}
      \textbf{Key arguments:}
      \begin{description}\small
        \item[\texttt{batch\_first=True}] Input shape: \texttt{(batch, seq\_len, features)}
        \item[\texttt{num\_layers}] Stacked LSTMs (2--3 common)
        \item[\texttt{dropout}] Applied between LSTM layers
        \item[\texttt{h[-1]}] Last layer's hidden state at final time step
              --- encodes the full sequence
      \end{description}
      \vspace{0.1cm}
      \begin{keybox}
        \small Typical RSXFS result: LSTM RMSE $\approx$ 1,400--1,600
        with sequence length 24 (2 years of history).
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Attention and Transformers}
% =============================================================================

\sectionslide{Attention and Transformers}{%
  Instead of reading sequences step-by-step,
  attention lets the model look at all positions simultaneously
  and decide which ones matter.}

% --- Slide: Attention mechanism ----------------------------------------------
\begin{frame}{The Attention Mechanism}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Limitation of LSTMs:} even with gating,
      information from 50+ steps ago may be lost or diluted.

      \vspace{0.1cm}
      \textbf{Attention idea:} at each output step,
      let the model directly attend to \emph{any} input step,
      weighted by relevance.

      \begin{definitionbox}{Scaled Dot-Product Attention \parencite{Vaswani2017}}
        \[
          \text{Attention}(Q,K,V) = \text{softmax}\!\left(\frac{QK^\top}{\sqrt{d_k}}\right)V
        \]
        $Q$ = queries, $K$ = keys, $V$ = values; $d_k$ = key dimension.
      \end{definitionbox}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{examplebox}{Business analogy}
        \small
        Predicting December sales: the model assigns high
        attention weight to last December (lag-12) and the
        preceding November, and low weight to random months
        in between. The attention score is \emph{learned},
        not hard-coded.
      \end{examplebox}
      \vspace{0.1cm}
      \begin{keybox}
        \small Attention computes a \textbf{weighted average}
        of the values $V$, where weights come from how
        well queries $Q$ match keys $K$.
        High similarity $\to$ high weight.
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: Transformer architecture ----------------------------------------
\begin{frame}{The Transformer Architecture}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Key components of a Transformer encoder block:}
      \begin{enumerate}\small
        \item \textbf{Multi-head self-attention:}
              run $h$ attention heads in parallel, each attending
              to different aspects of the sequence.
              Concatenate and project: $\text{MultiHead}(Q,K,V) = \text{concat}(\text{head}_1,\ldots,\text{head}_h)W^O$
        \item \textbf{Residual connection + LayerNorm:}
              $\mathbf{z} = \text{LayerNorm}(\mathbf{x} + \text{MultiHead}(\mathbf{x}))$
        \item \textbf{Feed-forward sublayer:}
              position-wise FFN applied to each position independently.
        \item \textbf{Positional encoding:}
              add sinusoidal position embeddings so the model
              knows token order (attention itself is permutation-invariant).
      \end{enumerate}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{keybox}
        \small
        \textbf{Why Transformers dominate NLP:}
        \begin{itemize}\small
          \item No sequential bottleneck --- all positions processed
                in parallel (vs.\ LSTM's sequential steps)
          \item Explicit long-range dependencies via attention weights
          \item Scales to billions of parameters (BERT, GPT)
        \end{itemize}
      \end{keybox}
      \vspace{0.1cm}
      \begin{examplebox}{BERT and GPT}
        \small BERT \parencite{Devlin2019} (encoder-only) for
        classification and embedding.
        GPT (decoder-only) for next-token generation.
        Both are Transformer variants pre-trained on
        massive text corpora.
      \end{examplebox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: Transformer for time series -------------------------------------
\begin{frame}{Transformer for Time Series Forecasting}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \begin{examplebox}{Minimal Transformer Encoder in PyTorch}
        {\small
        \texttt{encoder\_layer = nn.TransformerEncoderLayer(}\\
        \texttt{  d\_model=64,}\\
        \texttt{  nhead=4,}\\
        \texttt{  dim\_feedforward=256,}\\
        \texttt{  dropout=0.1,}\\
        \texttt{  batch\_first=True}\\
        \texttt{)}\\
        \texttt{encoder = nn.TransformerEncoder(}\\
        \texttt{  encoder\_layer, num\_layers=2)}\\[4pt]
        \texttt{class TSTransformer(nn.Module):}\\
        \texttt{  def \_\_init\_\_(self, in\_dim):}\\
        \texttt{    super().\_\_init\_\_()}\\
        \texttt{    self.proj = nn.Linear(in\_dim, 64)}\\
        \texttt{    self.enc  = encoder}\\
        \texttt{    self.fc   = nn.Linear(64, 1)}\\
        \texttt{  def forward(self, x):}\\
        \texttt{    h = self.enc(self.proj(x))}\\
        \texttt{    return self.fc(h[:, -1]).squeeze(-1)}
        }
      \end{examplebox}
    \end{column}
    \begin{column}{0.44\textwidth}
      \textbf{Practical notes for time series:}
      \begin{itemize}\small
        \item Add \textbf{positional encoding} (or learnable position embeddings)
              so the model knows temporal order
        \item Use \textbf{causal masking} to prevent the model from attending
              to future positions during training
        \item Transformers tend to outperform LSTMs on
              \emph{longer} sequences ($T > 100$) and
              \emph{multi-variate} time series
      \end{itemize}
      \vspace{0.1cm}
      \begin{warningbox}
        \small Transformers need more data than LSTMs.
        For short monthly business series ($T < 200$),
        LSTM is often a better starting point.
      \end{warningbox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Key Takeaways}
% =============================================================================

\sectionslide{Key Takeaways}{%
  RNNs, LSTMs, and Transformers are the core sequence architectures
  in modern deep learning.}

% --- Slide: Architecture comparison -----------------------------------------
\begin{frame}{Sequence Architecture Comparison}
  \begin{center}
    {\small
    \begin{tabular}{lllll}
      \toprule
      \textbf{Architecture} & \textbf{Long-range} & \textbf{Parallel} & \textbf{Data need} & \textbf{Best for} \\
      \midrule
      Vanilla RNN & Poor & No & Low & Very short sequences \\
      LSTM & Good & No & Medium & Monthly/quarterly TS \\
      1D CNN & Medium & Yes & Medium & Short TS, fast training \\
      Transformer & Excellent & Yes & High & Long sequences, NLP, multivariate \\
      \bottomrule
    \end{tabular}
    }
  \end{center}
  \begin{keybox}
    \textbf{Practical recommendation:}
    \begin{itemize}\small
      \item Short monthly series ($T < 200$, univariate): start with LSTM
      \item Long series or many features: try Transformer
      \item Fast prototyping: 1D CNN is quickest to train
      \item Production at scale: pre-trained Transformer (e.g., TimesFM, Chronos)
    \end{itemize}
  \end{keybox}
\end{frame}

% --- Slide: Key takeaways ---------------------------------------------------
\begin{frame}{Lecture 9: Key Takeaways}
  \begin{keybox}
    {\small
    \begin{enumerate}
      \item \textbf{RNNs} process sequences step-by-step, passing a hidden state $\mathbf{h}_t$
            across time steps. Vanilla RNNs suffer from the vanishing gradient problem
            and forget information after $\sim$10 steps.

      \item \textbf{LSTMs} add a cell state $\mathbf{c}_t$ and three gates (forget, input, output)
            that control information flow. The additive cell update stabilizes gradients
            over long sequences.

      \item \textbf{Attention} computes a weighted average over all positions, where weights
            reflect query-key similarity. It gives direct access to any past time step.

      \item \textbf{Transformers} stack multi-head self-attention with residual connections
            and feed-forward layers. All positions are processed in parallel ---
            no sequential bottleneck.

      \item In PyTorch: use \texttt{nn.LSTM} for sequence models and
            \texttt{nn.TransformerEncoder} for Transformer encoders.
            Always use \texttt{batch\_first=True} for time series data.
    \end{enumerate}
    }
  \end{keybox}
  \muted{\itshape Next: Bayesian Statistics I --- foundations of probabilistic
  inference for business forecasting.}
\end{frame}

% --- References ---------------------------------------------------------------
\begin{frame}[allowframebreaks]{References}
  \printbibliography[heading=none]
\end{frame}

% =============================================================================
\end{document}
% =============================================================================
