% =============================================================================
% Lecture 11: Bayesian Statistics II — Time Series & Hierarchical Models
% BSAD 8310: Business Forecasting
% University of Nebraska at Omaha
% =============================================================================

\documentclass[aspectratio=169, 11pt]{beamer}

\input{header}

% ---- Lecture metadata --------------------------------------------------------
\title{Bayesian Statistics II: Time Series \& Hierarchical Models}
\subtitle{BSAD 8310: Business Forecasting --- Lecture 11}
\author{Department of Economics}
\institute{University of Nebraska at Omaha}
\date{Spring 2026}

% =============================================================================
\begin{document}
% =============================================================================

\begin{frame}
  \titlepage
\end{frame}

% ---- Outline ----------------------------------------------------------------
\begin{frame}{Lecture Outline}
  \tableofcontents
\end{frame}

% =============================================================================
\section{Bayesian Time Series: Trend and Seasonality}
% =============================================================================

\sectionslide{Bayesian Time Series: Trend and Seasonality}{%
  Encode structural knowledge about trend and seasonality directly
  in the model, with uncertainty estimates on each component.}

% --- Slide: Why Bayesian TS? ------------------------------------------------
\begin{frame}{From Frequentist to Bayesian Time Series}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{What ARIMA and ETS give you:}
      \begin{itemize}\small
        \item Point forecasts with asymptotic prediction intervals
        \item Implicit assumptions about trend and error structure
        \item Limited ability to incorporate domain knowledge
        \item Single-equation, single-series models
      \end{itemize}
      \vspace{0.2cm}
      \textbf{What Bayesian TS adds:}
      \begin{itemize}\small
        \item Full posterior distributions over trend, seasonality,
              and each forecast horizon
        \item Explicit priors to incorporate business knowledge
              (``growth rate is unlikely to exceed 20\%'')
        \item Natural uncertainty propagation through time
        \item Hierarchical structure for multiple related series
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{examplebox}{Business Application}
        A retail chain has 500 stores. Rather than fitting
        500 independent ARIMA models, a Bayesian hierarchical
        model shares information across stores:
        \begin{itemize}\small
          \item Each store gets its own parameters
          \item Parameters are drawn from a common
                distribution (the hierarchy)
          \item Sparse data stores borrow strength
                from stores with more data
        \end{itemize}
      \end{examplebox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: Bayesian local linear trend model --------------------------------
\begin{frame}{Bayesian Structural Time Series}
  We decompose a time series into three components: a slowly-changing
  \textbf{trend}, a repeating \textbf{seasonal pattern}, and random \textbf{noise}.
  Each component gets its own parameters --- and the Bayesian approach
  gives us uncertainty estimates on all of them simultaneously.
  \[
    y_t = \mu_t + s_t + \varepsilon_t, \quad \varepsilon_t \sim \mathcal{N}(0, \sigma^2)
  \]
  where the trend $\mu_t$ evolves according to:
  \[
    \mu_t = \mu_{t-1} + \delta_{t-1} + \eta_t, \qquad
    \delta_t = \delta_{t-1} + \zeta_t
  \]
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Seasonal component using Fourier features:}

      Seasonality is modeled as smooth waves (sine and cosine curves)
      that repeat every year. More waves ($K$) = more flexible shapes.
      \[
        s_t = \sum_{k=1}^{K} \left[a_k \cos\!\left(\frac{2\pi k t}{P}\right) + b_k \sin\!\left(\frac{2\pi k t}{P}\right)\right]
      \]
      where $P$ = period (12 for monthly, 4 for quarterly)
      and $K$ controls flexibility.
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{definitionbox}{Local Linear Trend}
        $\mu_t$: level at time $t$\\
        $\delta_t$: slope (trend growth rate)\\
        $\eta_t \sim \mathcal{N}(0, \sigma_\eta^2)$: level shock\\
        $\zeta_t \sim \mathcal{N}(0, \sigma_\zeta^2)$: slope shock\\[3pt]
        Priors on $\sigma_\eta, \sigma_\zeta$ control how much the
        trend is allowed to change over time.
      \end{definitionbox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: PyMC Bayesian TS -------------------------------------------------
\begin{frame}{Bayesian TS in PyMC}
  \begin{columns}[T]
    \begin{column}{0.55\textwidth}
      \begin{examplebox}{Linear Trend + Seasonality}
        {\small
        \texttt{with pm.Model() as ts\_model:}\\
        \texttt{  \# Trend priors}\\
        \texttt{  alpha = pm.Normal('alpha', mu=y.mean(), sigma=y.std())}\\
        \texttt{  beta  = pm.Normal('beta',  mu=0, sigma=1)}\\[4pt]
        \texttt{  \# Fourier seasonality (K=3)}\\
        \texttt{  K = 3; P = 12}\\
        \texttt{  t\_idx = np.arange(len(y))}\\
        \texttt{  fourier = [pm.Normal(f'a\{k\}', 0, 1) *}\\
        \texttt{             np.cos(2*np.pi*k*t\_idx/P) +}\\
        \texttt{             pm.Normal(f'b\{k\}', 0, 1) *}\\
        \texttt{             np.sin(2*np.pi*k*t\_idx/P)}\\
        \texttt{             for k in range(1, K+1)]}\\
        \texttt{  seasonality = sum(fourier)}\\[4pt]
        \texttt{  mu = alpha + beta * t\_idx + seasonality}\\
        \texttt{  sigma = pm.HalfNormal('sigma', sigma=y.std())}\\
        \texttt{  obs = pm.Normal('obs', mu=mu,}\\
        \texttt{                  sigma=sigma, observed=y)}\\
        \texttt{  trace = pm.sample(2000)}
        }
      \end{examplebox}
    \end{column}
    \begin{column}{0.41\textwidth}
      \textbf{What you get:}
      \begin{itemize}\small
        \item Posterior distribution over the slope $\beta$:
              is the trend growing or declining? With what probability?
        \item Posterior over Fourier coefficients $a_k, b_k$:
              how strong and what shape is the seasonality?
        \item Posterior predictive distribution: credible bands
              around each forecast horizon, not just a point estimate
      \end{itemize}
      \vspace{0.1cm}
      \begin{keybox}
        \small Use \texttt{pm.sample\_posterior\_predictive(trace)}
        to generate forecast draws.
        Summarize as: median forecast + 90\% credible band.
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Hierarchical Models: Partial Pooling}
% =============================================================================

\sectionslide{Hierarchical Models: Partial Pooling}{%
  Share information across groups via a common prior distribution.
  The hierarchy learns what is typical; each group learns its own deviation.}

% --- Slide: The pooling problem ----------------------------------------------
\begin{frame}{Three Approaches to Multi-Group Data}
  Suppose we have sales data for $J = 50$ stores, each with limited history.
  \begin{columns}[T]
    \begin{column}{0.33\textwidth}
      \textbf{Complete pooling:}
      Fit one model ignoring stores.
      {\setbeamertemplate{itemize item}{\negc{$\times$}}
      \begin{itemize}\small
        \item Assumes all stores are identical
        \item Misses systematic store differences
        \item Wrong predictions for outlier stores
      \end{itemize}}
    \end{column}
    \begin{column}{0.33\textwidth}
      \textbf{No pooling:}
      Fit 50 independent models.
      {\setbeamertemplate{itemize item}{\negc{$\times$}}
      \begin{itemize}\small
        \item Ignores shared structure
        \item Noisy estimates for stores with few observations
        \item Overfits sparse data stores
      \end{itemize}}
    \end{column}
    \begin{column}{0.33\textwidth}
      \textbf{Partial pooling (hierarchical):}
      Store params drawn from a common prior.
      {\setbeamertemplate{itemize item}{\pos{$\checkmark$}}
      \begin{itemize}\small
        \item Each store has its own estimate
        \item Sparse stores shrink toward the group mean
        \item Data-rich stores less affected by prior
      \end{itemize}}
    \end{column}
  \end{columns}
  \vspace{0.2cm}
  \begin{keybox}
    Partial pooling is the Bayesian answer to the bias-variance trade-off:
    adding a hierarchical prior reduces variance (less overfitting) at the cost
    of a small bias (shrinkage toward the mean).
    \parencite{Gelman2013}
  \end{keybox}
\end{frame}

% --- Slide: Hierarchical model -----------------------------------------------
\begin{frame}{Hierarchical Model: Structure and Code}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Partial pooling for store-level intercepts:}
      \begin{align*}
        y_{it} &\sim \mathcal{N}(\alpha_i + \beta t, \sigma^2) \\
        \alpha_i &\sim \mathcal{N}(\mu_\alpha, \sigma_\alpha^2) \quad \text{(store hierarchy)} \\
        \mu_\alpha &\sim \mathcal{N}(\bar{y}, \text{SD}(y)) \quad \text{(group prior)} \\
        \sigma_\alpha &\sim \text{HalfNormal}(1)
      \end{align*}

      \begin{examplebox}{PyMC Partial Pooling}
        {\small
        \texttt{with pm.Model() as hier:}\\
        \texttt{  mu\_a  = pm.Normal('mu\_a', 0, 1)}\\
        \texttt{  sig\_a = pm.HalfNormal('sig\_a', 1)}\\
        \texttt{  alpha = pm.Normal('alpha',}\\
        \texttt{    mu=mu\_a, sigma=sig\_a, shape=J)}\\
        \texttt{  beta = pm.Normal('beta', 0, 1)}\\
        \texttt{  mu = alpha[store\_idx] + beta * t}\\
        \texttt{  pm.Normal('obs', mu=mu,}\\
        \texttt{    sigma=1, observed=y)}
        }
      \end{examplebox}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{keybox}
        \textbf{The key line:}
        \texttt{alpha = pm.Normal('alpha', mu=mu\_a, sigma=sig\_a, shape=J)}

        Each store $i$ gets its own $\alpha_i$, but all $\alpha_i$ are
        drawn from $\mathcal{N}(\mu_\alpha, \sigma_\alpha)$.
        Learning $\mu_\alpha$ and $\sigma_\alpha$ from data is what
        makes it hierarchical.
      \end{keybox}
      \vspace{0.1cm}
      \begin{warningbox}
        \small Use the \textbf{non-centered parameterization}
        for better sampling efficiency:
        $\alpha_i = \mu_\alpha + \sigma_\alpha \cdot z_i$,
        $z_i \sim \mathcal{N}(0,1)$.
        PyMC tutorials demonstrate this.
      \end{warningbox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Key Takeaways}
% =============================================================================

\sectionslide{Key Takeaways}{%
  Bayesian TS and hierarchical models: full uncertainty,
  shared structure, better predictions for sparse groups.}

% --- Slide: Key takeaways ---------------------------------------------------
\begin{frame}{Lecture 11: Key Takeaways}
  \begin{keybox}
    {\small
    \begin{enumerate}
      \item Bayesian time series models decompose the series into
            \textbf{trend}, \textbf{seasonality}, and \textbf{error},
            each with its own prior. The posterior quantifies uncertainty
            in every component simultaneously.

      \item \textbf{Fourier features} ($\cos$ and $\sin$ terms) parameterize
            seasonality with $K$ harmonics. More harmonics = more flexible
            seasonal patterns.

      \item \textbf{Hierarchical models} use partial pooling to share
            information across groups (stores, regions, products).
            Sparse groups shrink toward the group mean; data-rich groups
            are less affected.

      \item The three pooling options --- complete, no, and partial ---
            represent a bias-variance trade-off. Partial pooling
            (hierarchical) is almost always the best choice for
            multi-group business data.

      \item In PyMC, hierarchical structure is created by giving
            group-level parameters a shared prior with hyperparameters
            that are themselves estimated from the data.
    \end{enumerate}
    }
  \end{keybox}
  \muted{\itshape Next: Bayesian Statistics III --- Bayesian linear regression,
  posterior coefficient distributions, and conditional analysis.}
\end{frame}

% --- References ---------------------------------------------------------------
\begin{frame}[allowframebreaks]{References}
  \printbibliography[heading=none]
\end{frame}

% =============================================================================
\end{document}
% =============================================================================
