% =============================================================================
% Lecture 08: CNN Architectures
% BSAD 8310: Business Forecasting
% University of Nebraska at Omaha
% =============================================================================

\documentclass[aspectratio=169, 11pt]{beamer}

\input{header}

% ---- Lecture metadata --------------------------------------------------------
\title{CNN Architectures}
\subtitle{BSAD 8310: Business Forecasting --- Lecture 8}
\author{Department of Economics}
\institute{University of Nebraska at Omaha}

% =============================================================================
\begin{document}
% =============================================================================

\begin{frame}
  \titlepage
\end{frame}

% ---- Outline ----------------------------------------------------------------
\begin{frame}{Lecture Outline}
  \tableofcontents
\end{frame}

% =============================================================================
\section{Convolution: The Core Idea}
% =============================================================================

\sectionslide{Convolution: The Core Idea}{%
  Instead of connecting every neuron to every input,
  a convolutional filter scans locally and shares weights.}

% --- Slide: Motivation -------------------------------------------------------
\begin{frame}{Why Convolution?}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Problem with fully connected layers:}
      \begin{itemize}\small
        \item For a $224 \times 224$ image, the input has 50,176 pixels.
              A single hidden layer with 1,000 neurons needs
              $50{,}176 \times 1{,}000 = 50\,\text{M}$ weights --- just for one layer.
        \item Ignores spatial structure: nearby pixels are related;
              distant pixels usually are not.
        \item Does not generalize: a pattern learned in the top-left
              corner is not reused elsewhere.
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{keybox}
        \textbf{Convolutional layer solution:}
        \begin{itemize}\small
          \item \textbf{Local connectivity:} each neuron sees only a small
                \textbf{receptive field} (e.g., $3 \times 3$ pixels)
          \item \textbf{Weight sharing:} the same filter is applied
                everywhere in the image
          \item \textbf{Result:} far fewer parameters; translation invariance
        \end{itemize}
      \end{keybox}
    \end{column}
  \end{columns}
  \vspace{0.2cm}
  \begin{examplebox}{Analogy: Scanning a document}
    A human reader does not memorize a font separately for each position
    on the page. They learn to recognize the letter ``A'' once, then
    apply that recognition wherever it appears. Convolutional filters
    do exactly this.
  \end{examplebox}
\end{frame}

% --- Slide: How convolution works --------------------------------------------
\begin{frame}{How a Convolutional Layer Works}
  Imagine sliding a small magnifying glass along a sequence.
  At each position, the glass examines a short window of values and
  computes a single number indicating how strongly a pattern is present.
  This is exactly what a convolutional filter does.
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{1D convolution (for sequences):}
      \[
        (f * g)[t] = \sum_{k=0}^{K-1} f[k] \cdot g[t - k]
      \]
      Slide a filter of length $K$ across the sequence,
      computing a dot product (a weighted sum) at each position.

      \vspace{0.2cm}
      \textbf{Key parameters:}
      \begin{description}\small
        \item[Kernel size] Length of the filter ($K$). Larger = wider receptive field.
        \item[Stride] Steps between positions. Stride 2 halves the output length.
        \item[Padding] Add zeros at edges to keep output the same length
              (\texttt{padding='same'}).
        \item[Filters] Number of different filters to learn (output channels).
      \end{description}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{definitionbox}{What does a filter detect?}
        Each filter learns to detect a specific \textbf{local pattern}
        (e.g., edges, spikes, seasonal dips).
        The output of a filter is called a \textbf{feature map}:
        high values indicate where the pattern was found.
      \end{definitionbox}
      \vspace{0.1cm}
      \begin{keybox}
        \small After the convolution, apply an activation (ReLU) and
        then \textbf{pooling} to reduce the feature map size.
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Pooling and Dropout}
% =============================================================================

\sectionslide{Pooling and Dropout}{%
  Pooling reduces spatial size and adds robustness.
  Dropout prevents co-adaptation of filters.}

% --- Slide: Pooling ----------------------------------------------------------
\begin{frame}{Pooling: Downsampling Feature Maps}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      After a convolutional layer, apply \textbf{pooling}
      to reduce the size of each feature map:
      \begin{itemize}\small
        \item \textbf{Max pooling:} take the maximum value in each
              non-overlapping window. Retains the most prominent
              activation --- good for detecting if a feature is
              \emph{present anywhere} in the window.
        \item \textbf{Average pooling:} take the mean. Smoother,
              preserves intensity information.
        \item \textbf{Global average pooling:} collapse the entire
              feature map to a single number. Commonly used before
              the final fully connected layer.
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{examplebox}{Pooling in PyTorch}
        {\small
        \texttt{nn.MaxPool1d(kernel\_size=2)}\\
        \texttt{nn.AvgPool1d(kernel\_size=2)}\\
        \texttt{nn.AdaptiveAvgPool1d(1)}
        \hfill \textit{(global)}
        }
      \end{examplebox}
      \vspace{0.2cm}
      \begin{keybox}
        \small Pooling provides \textbf{translation invariance}:
        a pattern detected slightly earlier or later in a sequence
        still produces the same pooled output.
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{CNN Architecture Evolution}
% =============================================================================

\sectionslide{CNN Architecture Evolution}{%
  LeNet $\to$ VGG $\to$ Inception $\to$ ResNet:
  each generation solved a specific limitation.}

\begin{frame}{}
  \begin{keybox}
    These architectures were developed for images, but the same design
    principles --- stacking small filters, skip connections, efficient depth ---
    carry directly into 1D CNNs for time series.
    Understanding the history helps you make better architectural choices.
  \end{keybox}
\end{frame}

% --- Slide: LeNet and VGG ----------------------------------------------------
\begin{frame}{LeNet to VGG: Depth and Simplicity}
  \begin{columns}[T]
    \begin{column}{0.50\textwidth}
      \textbf{LeNet-5 \parencite{LeCun1998}:}
      \begin{itemize}\small
        \item First practical CNN (1998)
        \item 2 conv layers + 3 FC layers; ~60K parameters
        \item Input: $32 \times 32$ grayscale; Output: 10 digit classes
        \item Proved that learned filters beat hand-crafted features
      \end{itemize}
      \vspace{0.1cm}
      \textbf{VGG-16/19 \parencite{Simonyan2015}:}
      \begin{itemize}\small
        \item 16--19 layers, all using $3 \times 3$ kernels
        \item Key insight: stack multiple $3\times3$ filters instead of
              one large filter --- same receptive field, fewer parameters
        \item 138M parameters; very deep but simple pattern
        \item Still a strong baseline for transfer learning
      \end{itemize}
    \end{column}
    \begin{column}{0.46\textwidth}
      \begin{keybox}
        \textbf{The $3\times3$ rule:}
        Two $3\times3$ conv layers have the same receptive field
        as one $5\times5$ layer, but use
        $2 \times (3^2) = 18$ vs.\ $5^2 = 25$ parameters per channel.
        Three $3\times3$ layers match a $7\times7$ layer
        at 27 vs.\ 49 parameters.

        Stack small filters, go deep.
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: Inception and ResNet --------------------------------------------
\begin{frame}{Inception and ResNet: Efficient Depth}
  \begin{columns}[T]
    \begin{column}{0.50\textwidth}
      \textbf{GoogLeNet / Inception \parencite{Szegedy2015}:}
      \begin{itemize}\small
        \item Inception module: apply $1\times1$, $3\times3$, $5\times5$
              filters \textbf{in parallel}, then concatenate outputs
        \item $1\times1$ convolutions reduce channel depth before expensive
              large-kernel operations (bottleneck)
        \item 22 layers deep; only 5M parameters vs.\ VGG's 138M
      \end{itemize}
      \vspace{0.1cm}
      \textbf{ResNet \parencite{He2016}:}
      \begin{itemize}\small
        \item \textbf{Residual connections:}
              $\mathbf{a}^{(\ell+2)} = \mathcal{F}(\mathbf{a}^{(\ell)}) + \mathbf{a}^{(\ell)}$
        \item Gradient flows directly through the skip connection ---
              solves the vanishing gradient problem in very deep nets
        \item Enables 50, 101, 152-layer networks that still train stably
      \end{itemize}
    \end{column}
    \begin{column}{0.46\textwidth}
      \begin{definitionbox}{Residual (Skip) Connection}
        \[
          \mathbf{a}^{(\ell+2)} = \underbrace{\mathcal{F}(\mathbf{a}^{(\ell)})}_{\text{learned residual}} + \underbrace{\mathbf{a}^{(\ell)}}_{\text{identity}}
        \]
        If the layer learns nothing useful, it can set
        $\mathcal{F} \approx 0$ and pass the input unchanged.
        The network only needs to learn the \emph{residual}.
      \end{definitionbox}
      \begin{keybox}
        \small ResNet skip connections are now standard
        across deep learning: Transformers, U-Nets, and
        most modern architectures include them.
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{1D CNNs for Time Series}
% =============================================================================

\sectionslide{1D CNNs for Time Series}{%
  Apply convolutional filters along the time axis
  to detect local temporal patterns automatically.}

% --- Slide: 1D CNN motivation ------------------------------------------------
\begin{frame}{Why Use a 1D CNN for Forecasting?}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      A \textbf{1D CNN} applies filters along the \emph{time} dimension:
      \begin{itemize}\small
        \item Each filter learns to detect a local temporal pattern
              (a spike, a dip, a ramp-up over $K$ consecutive steps)
        \item Multiple filters in parallel learn different patterns
        \item Stacking layers increases the effective look-back window:
              2 layers of kernel size 5 see 9 time steps in total
      \end{itemize}
      \vspace{0.2cm}
      \textbf{Compared to lag features + XGBoost:}
      \begin{itemize}\small
        \item No manual feature engineering needed
        \item Patterns are learned, not hand-coded
        \item But: needs more data to learn effectively
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \begin{examplebox}{1D CNN in PyTorch}
        {\small
        \texttt{nn.Conv1d(}\\
        \texttt{  in\_channels=1,}\\
        \texttt{  out\_channels=32,}\\
        \texttt{  kernel\_size=5,}\\
        \texttt{  padding=2}\\
        \texttt{)}
        }
        \vspace{0.1cm}
        {\footnotesize
        \textbf{Input shape:} \texttt{(batch, channels, seq\_len)}\\
        A univariate series: \texttt{in\_channels=1}\\
        Multiple features: \texttt{in\_channels=p}
        }
      \end{examplebox}
      \begin{keybox}
        \small 1D CNNs are fast to train and
        often competitive with LSTMs on
        shorter sequences (< 500 time steps).
      \end{keybox}
    \end{column}
  \end{columns}
\end{frame}

% --- Slide: Full CNN forecasting pipeline ------------------------------------
\begin{frame}{1D CNN Forecasting Pipeline in PyTorch}
  \begin{columns}[T]
    \begin{column}{0.55\textwidth}
      \begin{examplebox}{CNN Regressor for Time Series}
        {\small
        \texttt{class CNNForecaster(nn.Module):}\\
        \texttt{  def \_\_init\_\_(self, seq\_len, n\_feat):}\\
        \texttt{    super().\_\_init\_\_()}\\
        \texttt{    self.conv = nn.Sequential(}\\
        \texttt{      nn.Conv1d(n\_feat, 32, kernel\_size=5, padding=2),}\\
        \texttt{      nn.ReLU(),}\\
        \texttt{      nn.Conv1d(32, 64, kernel\_size=3, padding=1),}\\
        \texttt{      nn.ReLU(),}\\
        \texttt{      nn.AdaptiveAvgPool1d(1)}\\
        \texttt{    )}\\
        \texttt{    self.fc = nn.Linear(64, 1)}\\[4pt]
        \texttt{  def forward(self, x):}\\
        \texttt{    \# x shape: (batch, n\_feat, seq\_len)}\\
        \texttt{    h = self.conv(x).squeeze(-1)}\\
        \texttt{    return self.fc(h).squeeze(-1)}
        }
      \end{examplebox}
    \end{column}
    \begin{column}{0.41\textwidth}
      \textbf{Architecture walkthrough:}
      \begin{enumerate}\small
        \item \texttt{Conv1d(n\_feat, 32, 5)}: 32 filters of width 5 applied to all features
        \item \texttt{ReLU}: non-linearity
        \item \texttt{Conv1d(32, 64, 3)}: 64 filters of width 3 applied to 32 feature maps
        \item \texttt{AdaptiveAvgPool1d(1)}: collapse to one value per channel (global average)
        \item \texttt{Linear(64, 1)}: single forecast output
      \end{enumerate}
      \vspace{0.1cm}
      \muted{\small Typical RSXFS result: 1D CNN RMSE $\approx$ 1,600--1,800 with proper windowing.}
    \end{column}
  \end{columns}
\end{frame}

% =============================================================================
\section{Key Takeaways}
% =============================================================================

\sectionslide{Key Takeaways}{%
  CNN architectures: from local filters to deep residual networks.}

% --- Slide: Architecture summary --------------------------------------------
\begin{frame}{CNN Architecture Summary}
  \begin{center}
    {\small
    \begin{tabular}{llll}
      \toprule
      \textbf{Architecture} & \textbf{Year} & \textbf{Key Innovation} & \textbf{Params} \\
      \midrule
      LeNet & 1998 & First practical CNN; conv + pool & $\sim$60K \\
      VGG & 2015 & Deep stack of $3\times3$ filters & $\sim$138M \\
      Inception & 2015 & Parallel multi-scale filters; $1\times1$ bottleneck & $\sim$5M \\
      ResNet & 2016 & Residual (skip) connections; train 100+ layers & $\sim$25M \\
      \bottomrule
    \end{tabular}
    }
  \end{center}
  \begin{keybox}
    \textbf{For business forecasting with 1D CNNs:}
    \begin{itemize}\small
      \item Use 2--3 \texttt{Conv1d} layers with kernel size 3--7
      \item Follow with ReLU and global average pooling
      \item Skip connections help if stacking more than 4 layers
      \item Faster to train than LSTMs; competitive accuracy on shorter sequences
    \end{itemize}
  \end{keybox}
\end{frame}

% --- Slide: Key takeaways ---------------------------------------------------
\begin{frame}{Lecture 8: Key Takeaways}
  \begin{keybox}
    {\small
    \begin{enumerate}
      \item \textbf{Convolutional layers} apply learned filters locally across input
            positions. This gives \textbf{weight sharing} (same filter reused everywhere)
            and \textbf{translation invariance} (pattern recognized regardless of position).

      \item \textbf{Pooling} downsamples feature maps. Max pooling retains peak
            activations; global average pooling collapses each feature map to
            a single value before the output layer.

      \item The $3\times3$ rule: stack small filters (VGG principle).
            Two $3\times3$ layers cover the same field as one $5\times5$
            with fewer parameters.

      \item \textbf{Residual connections} (ResNet) add the input directly to the
            layer output: $\mathbf{a}^{(\ell+2)} = \mathcal{F}(\mathbf{a}^{(\ell)}) + \mathbf{a}^{(\ell)}$.
            This stabilizes gradients in very deep networks.

      \item \textbf{1D CNNs} apply filters along the time axis to detect local
            temporal patterns. Faster to train than LSTMs; a strong
            alternative for shorter sequences.
    \end{enumerate}
    }
  \end{keybox}
  \muted{\itshape Next: RNNs, LSTMs \& Transformers ---
  architectures with explicit memory and attention over sequences.}
\end{frame}

% --- References ---------------------------------------------------------------
\begin{frame}[allowframebreaks]{References}
  \printbibliography[heading=none]
\end{frame}

% =============================================================================
\end{document}
% =============================================================================
